Comparison of Fast Imaging Architectures for Large, Multi-scale, Hierarchical Aperture Arrays

### 2024-06-14 ms.tex
Update on Overleaf.

The same reasoning can be extended for inter-station processing, where there is significant diversity. There is a large range in the number of stations and in the array filling factor. As a result, both the layout and cadence interval play significant roles in determining the efficiency of architectures. For arrays with small numbers of stations such as \texttt{LAMBDA-I} and \texttt{CASPA}, the XBF architecture holds an advantage on all timescales. For arrays with large numbers of stations such as \texttt{SKA1-low-core} with a reasonably high packing density, EPIC holds the advantage on all timescales, but with slower cadence intervals, $t_\textrm{acc}\gtrsim 100$~ms, XBF and XFFT also have comparable efficiency. Although the \texttt{SKA1-low} has a large number of stations, they are not as densely packed as in the core, and therefore, the role of $t_\textrm{acc}$ is more significant. XFFT and EPIC are comparable at $t_\textrm{acc}\lesssim 1$~ms, but for $t_\textrm{acc}\gtrsim 100$~ms, XFFT is clearly advantageous. Similarly, \texttt{FarView} has a similar array filling factor as \texttt{SKA1-low-core}, but has far fewer stations. Hence, EPIC is only advantageous for $t_\textrm{acc}\lesssim 0.1$~ms, whereas XFFT becomes more efficient for $t_\textrm{acc}\gtrsim 1$~ms.

@@

Besides computational costs, there are other practical considerations in the deployment of an architecture such as the hardware's processing bandwidth, internal memory bandwidth management, input and output data rates, etc. For example, the GPU-based deployment of EPIC on the LWA is severely limited by the memory bandwidth constraints on chip\footnote{See \url{https://github.com/epic-astronomy/Memos/blob/master/PDFs/009_EPIC_Code_Optimizations.md} for a detailed analysis.}. The architecture of current generation of GPUs is found to be inefficient at handling the transpose operation required in an $N$-point two-dimensional FFT, whereas they may be more efficient at handling correlations. On the contrary, the current generation of FPGAs seem more suited for low bit width two-dimensional FFT from computational and on-chip memory bandwidth perspectives. It is extremely difficult to forecast the continually evolving capabilities of hardware accessible to instruments at the instant of deployment. Thus, we restrict our analysis to the metric computational cost density, which is independent of evolving hardware capabilities. For example, even if the computations are parallelised over a large number of GPU cores as noted in \citet{Sokolowski+2024}, the total computational cost remains unaffected.

@@

>>>>

The same reasoning can be extended for inter-station processing, where there is significant diversity. There is a large range in the number of stations and in the array filling factor. As a result, both the layout and cadence interval play significant roles in determining the efficiency of architectures. For arrays with small numbers of stations such as \texttt{LAMBDA-I} and \texttt{CASPA}, the XBF architecture holds an advantage on all timescales. For arrays with large numbers of stations such as \texttt{SKA1-low-core} with a reasonably high packing density, EPIC holds the advantage on all timescales, but with slower cadence intervals, $t_\textrm{acc}\gtrsim 100$~ms, XBF and XFFT also have comparable efficiency. Although the \texttt{SKA1-low} has a large number of stations, they are not as densely packed as in the core, and therefore, the role of $t_\textrm{acc}$ is more significant. XFFT and EPIC are comparable at $t_\textrm{acc}\lesssim 1$~ms, but for $t_\textrm{acc}\gtrsim 100$~ms, XFFT is clearly advantageous. Similarly, \texttt{FarView} has a comparable array filling factor as \texttt{SKA1-low-core}, but has far fewer stations. Hence, EPIC is only advantageous for $t_\textrm{acc}\lesssim 0.1$~ms, whereas XFFT becomes more efficient for $t_\textrm{acc}\gtrsim 1$~ms.

@@

Besides computational costs, there are other practical considerations in the deployment of an architecture such as the hardware's processing bandwidth, internal memory bandwidth management, input and output data rates, etc. For example, the GPU-based deployment of EPIC on the LWA is severely limited by the memory bandwidth constraints on chip\footnote{See \url{https://github.com/epic-astronomy/Memos/blob/master/PDFs/009_EPIC_Code_Optimizations.md} for a detailed analysis.}. The architecture of current generation of GPUs is found to be inefficient at handling the transpose operation required in an $N$-point two-dimensional FFT, whereas they may be more efficient at handling correlations. On the contrary, the current generation of FPGAs seem more suited for low bit width two-dimensional FFT from computational and on-chip memory bandwidth perspectives. It is extremely difficult to forecast the continually evolving landscape of hardware capabilities accessible to instruments at the instant of deployment. Thus, we restrict our analysis to the metric of computational cost density, which is independent of evolving hardware capabilities. For example, even if the computations are parallelised over a large number of GPU cores as noted in \citet{Sokolowski+2024}, the total computational cost remains unaffected.

@@

Two high impact potential science drivers in radio astronomy are explosive transient phenomena and observational cosmology. The paradigm of aperture arrays used in radio astronomy is undergoing a transformation towards large numbers of low-cost elements grouped together in a hierarchy of spatial scales to achieve the desired science outcomes. Such a paradigm comes along with enormous challenges in computing and throughput because of the large numbers of elements spanning a multiplicity of spatial scales, and range of cadence intervals, real-time processing, and continuous operation. Different processing architectures are known to be optimal for specific layouts and processing cadences. However, a single architectural may not be optimal for a hierarchical spatial layout.

This paper explores hybrid architecture solutions for existing and planned multi-scale aperture arrays (\texttt{SKA1-low-core}, \texttt{SKA1-low}, \texttt{LAMBDA-I}, \texttt{CASPA}, and \texttt{FarView}) on a wide range of cadence intervals, 0.1~ms$\le t_\textrm{acc} \le 10$~s over the full field of view, using the metric of computational cost density evaluated over the discovery parameter space. The interferometric data processing architectures considered here are pre-correlation voltage beamformer (BF), a generic FFT-based direct imager (EPIC), beamforming from correlations (XBF), and FFT-based imaging of correlations (XFFT). The computational cost budget for various architecture combinations for these different arrays are broken down into their detailed components.

For densely packed layout with large numbers of elements, EPIC is computationally most efficient regardless of processing cadence. This was the case for the station layouts of all the arrays considered, and hence EPIC emerges as the computationally most efficient architecture for full field of view, station level data processing on all cadence intervals except for \texttt{CASPA} on $t_\textrm{acc}\gtrsim 100$~ms, when XFFT becomes more efficient. Thus, in almost cases of station layouts considered here, EPIC appears to hold a clear advantage for full field of view, intra-station processing.

The inter-station array layout has a wider range in the number of stations and sparsity of their distribution. In such cases, the cadence interval plays a significant role in determining the computationally most efficient architecture. For inter-station data processing in \texttt{LAMBDA-I} and \texttt{CASPA}, we find XBF to be the most suitable option for all cadence intervals, while for \texttt{SKA1-low-core} it is EPIC that holds the advantage on all cadence intervals. For \texttt{SKA1-low}, XFFT appears most efficient for $t_\textrm{acc}\gtrsim 1$~ms while EPIC and XFFT become comparable for $t_\textrm{acc}\lesssim 0.1$~ms. Thus, a multi-scale aperture array requires an optimal hybrid architecture strategy depending on the processing cadence. This study provides a guide for designing strategic data processing architecture for multi-scale aperture arrays, and conversely, can be used to design array parameters for specified computational complexity.

We have considered only computational cost density over the discovery parameter space as a metric for finding the most efficient combination of processing architectures for multi-scale aperture arrays. Additional computing will be required for post-processing steps such as de-dispersion, deconvolution, etc. depending on the application. Actual deployment will have to take other significant practical factors into account such as power consumption, memory bandwidth, data rate, cost and capabilities of current hardware.
